On the Gravitational stability of gravito-turbulent accretion disks

### 2016-01-26 paper/paper3d.tex
KMK edits

%\documentclass[12pt]{aastex}
\documentclass[iop, numberedappendix]{emulateapj}
%\documentclass{article}
%\usepackage{draftwatermark}
%\usepackage{paralist}
\shorttitle{Non-ideal Gravitational instability}
\shortauthors{M.-K.\ Lin \& K. Kratter}

@@

\begin{abstract}
Low mass, self-gravitating accretion disks admit quasi-steady,
`gravito-turbulent' states in which cooling balances turbulent
heating. However, numerical simulations show that gravito-turbulence
cannot be sustained beyond dynamical timescales when the cooling rate
or turbulent viscosity is too large, resulting in disk fragmentation.        
We motivate and quantify an interpretation of disk % a physical
                                % interpretation 
fragmentation as the inability to maintain gravito-turbulence due to  
secondary instabilities driven by:  
\begin{inparaenum}[1)] 
\item 
  cooling, which reduces pressure support; and/or
\item 
  viscosity, which reduces rotational support. 
\end{inparaenum}
We analyze the gravitational stability of accretion disks 
including these non-ideal effects explicitly.    
We derive a general dispersion relation for linear waves in a 
self-gravitating, viscous, non-adiabatic disk with internal heating,
external irradiation, and cooling. We  
consider parameterized cooling functions in two and three dimensions,
as well as full radiative diffusion in 3D. We show that 
%cooling/viscosity can destabilize the disk even when the classic
%Toomre stability condition is satisfied; 
generally there is no critical cooling rate/viscosity below which 
the disk is formally stable. 
We apply linear theory to 
global models of protoplanetary disks (PPDs) subject to
gravito-turbulence modeled as an effective 
viscosity and cooling via dust opacity. 
%We find the sole effect of
%cooling is insufficient to 
%destabilize a realistic PPD. 
We find that viscosity renders the PPD beyond $\sim 60$AU dynamically
unstable on radial %this is for Mdot = 1d-6  
lengthscales a few times the local disk thickness. This is coincident 
with the empirical condition for disk fragmentation based on a
maximum sustainable stress. 
We suggest turbulent stresses can play an active role in realistic disk 
fragmentation by removing rotational stabilization against 
self-gravity. 
\end{abstract}

\section{Introduction}\label{intro} 
Understanding the gravitational stability of rotating disks    
is central to many astrophysical problems. In the context of 
gaseous protoplanetary disks (PPDs), gravitational instability (GI)
has two applications. It can provide gravitational torques to
transport angular momentum outwards and thus enable mass accretion
\citep{armitage10,turner14}. GI may also lead to disk fragmentation,
which has been invoked to explain giant planet formation at large
radii \citep{helled14}. Studying these non-linear phenomena require
direct numerical simulations. Nevertheless, physical insight can be    
obtained through analytical modeling.

The standard metric for the (inverse) strength of disk self-gravity is
the \emph{Toomre parameter}, 
\begin{align}\label{toomreQ_criterion}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}  
\end{align}
\citep{toomre64}. Here, $c_s$ is the sound-speed,
$\kappa$ is the epicyclic frequency (which equals the rotation
frequency $\Omega$ in a Keplerian disk), $\Sigma$ is the surface
density and $G$ is the gravitational constant. The Toomre parameter is a
measure of the destabilizing effect of self-gravity ($G\Sigma$) against 
the stabilzing effect of rotation ($\kappa$) and pressure 
($c_s$). This is evident from the dispersion relation, 
\begin{align}\label{classic_dr}
  s^2 =  2\pi G\Sigma|k| - \kappa^2 - c_s^2k^2,   
\end{align}
which relates the growth rate $s$ and radial wavenumber $k$ for local,
axisymmetric waves in a two-dimensional (2D, razor-thin), inviscid and
adiabatic (including isothermal) disk. Such disturbances are formally
unstable when $Q<1$. Non-axisymmetric modes can develop for larger,
but still order-unity values of $Q$
\citep{papaloizou89,papaloizou91}. 
%strictly speaking non axi *local* modes are stable -> only transient
%growth
%true non-axi instability needs global disk  model 
%non-axi GI at slightly larger Q, but still characterized by Q

It is important to remember that the popular dispersion relation, 
Eq. \ref{classic_dr},  is derived from \emph{idealized} conditions: 
the disk is laminar and does not experience any net thermal
losses\footnote{Isothermal disks implicitly assume  
  heating and thermal losses are exactly balanced at all 
  times, so these do not cool in the current context.} 
(hereafter `cooling').  
However, real disks can cool, for example, %radiative, convective cooling
due to radiative losses. Accretion disks may also be turbulent, for
which its dynamic and thermodynamic effect is often modeled through an
effective viscosity \citep{shakura73,lin87,rafikov15}. {\bf 
  more refs here?} 
%in terms of AM transport and heating 
%assess GI in general

Despite this, the Toomre parameter and its associated
dispersion relation are still widely applied to viscous, cooling
accretion disk models \citep[e.g.][]{cossins09, kimura12}, {\bf more refs?}
even though Eq. \ref{toomreQ_criterion}---\ref{classic_dr} were 
derived from inviscid disks without cooling. 
%e.g. using classic DR to determine what scales need to be resolved

However, as we emphasize in this work, including non-ideal
physics, such as cooling and viscosity, modifies the classic dispersion
relation, and hence the condition for GI. Thus the Toomre condition
alone is insufficient to assess GI in realistic disks. The goal of
this work is to generalize the analytic treatment of disk GI to allow
a self-consistent discussion of GI in realistic accretion disk
models. Specifically, we will examine how  cooling and viscosity can
lead to GI even when $Q>1$, by removing pressure and rotational
support, respectively.

\subsection{Cooling-driven gravitational instability}\label{cool_gi}
Cooling reduces pressure support against self-gravity. The
cooling time  
$\tcool$ is the timescale over which the disk temperature $T$ is relaxed
to some floor value $T_0$, which may be zero. 
%That is, $\p_t T =
%-(T-T_0)/\tcool$. 
 It is often written as 
\begin{align} \label{beta_def}
  \tcool = \beta \Omega^{-1},
\end{align}
where $\beta$ is the corresponding dimensionless cooling time. This
type of parameterized cooling, first applied by \cite{gammie01},
allows the full range of thermodynamic response to be explored. 
{\bf physical origin of cooling in PPDs}

We will quantify how cooling enables GI
even when $Q>1$. For example, if perturbations are
cooled towards zero tempertaure, we find the above dispersion relation is
modified to read 
\begin{align}\label{cool_dr}
s^2 = 2\pi G\Sigma |k| - \kappa^2 - \gamma \left(\frac{\tcool s}{1 +
  \tcool s}\right)c_s^2k^2,
\end{align}
where $\gamma$ is the adiabatic index. Cooling increases the growth
rate by reducing the magnitude of the 
pressure term. In fact, for $t_c\in[0,\infty)$ a formal condition
for instability is  
\begin{align*}
  2\pi G \Sigma |k| > \kappa^2,
\end{align*}
which is what would be obtained from Eq. \ref{classic_dr} with 
pressure neglected ($c_s\to 0$). This condition does not actually
depend on the cooling rate, and instability is
possible for any finite $Q$. %at sufficiently small scales 
This shows that the mere presence of
cooling changes the qualitative nature of GI compared to the simple
Toomre condition.   
%{\bf numerical implications? if using standard beta cooling, then as
%  long as it's there then sufficiently small scales will be
 % unstbale. can access those scales with high resolution. defer to
 % discussion section}

@@

\subsection{Viscosity-driven gravitational instability}\label{visc_gi}
%there is also viscous damping which stabilizes
Viscous disks can also develop GI even when $Q>1$. This is because, as 
demonstrated below, viscosity eliminates rotational support against  
self-gravity for long-wavelength disturbances 
\citep{lynden-bell74,willerding92,gammie96}. A similar effect   
occurs in dusty fluids where the required frictional forces are 
provided by dust-gas drag \citep{ward00, takahashi14}. 
In fact, this
is a mechanism to enhance particle clumping for planetesimational
formation %frag of dust 
\citep{youdin11}. Then it is not unreasonable to expect an analogous
fragmentation of gaseous disks due to viscosity.

It is conventional to write the kinematic viscosity $\nu$ as   
\begin{align}\label{alpha_param}
  \nu  = \alpha \frac{c_s^2}{\Omega}, %dimensional argument 
\end{align}
where $\alpha$ is the dimensionless viscosity
coefficient. This
parameterization can be modified to include more complex dependencies on the
fluid variables, but Eq. \ref{alpha_param} is the general form. 
{\bf maybe no need to intro alpha param here?}

For an isothermal, viscous, self-gravitating disk in 2D,
\cite{gammie96} finds the approximate disperion relation   
\begin{align}
s \simeq \frac{\nu k^2\left(2\pi G\Sigma |k| -
  c_s^2k^2\right)}{\kappa^2 + c_s^2k^2 - 2\pi G\Sigma|k|}. 
\end{align}
Assuming $Q>1$, instability occurs if 
\begin{align*}
  2\pi G \Sigma > c_s^2|k|,
\end{align*}
provided that $\nu k^2\neq0$. This condition for viscous GI is 
identical that would be obtained from Eq. \ref{classic_dr} with
rotation neglected ($\Omega\to 0$). That is, a classically stable
disk can be destabilized by viscosity as it effectively removes
rotational stabilization \citep{lynden-bell74}.

%turbulence <-> viscosity assumption 
For gaseous accretion disks, a Navier-Stokes viscosity is often
implemented (as above) to mimic hydrodynamic or magneto-hydrodynamic
(MHD) turbulence \citep{shakura73}. How well turbulence can be modeled
as an effective viscosity is a separate issue 
\citep{balbus99}. However, it is reasonable to assume that turbulence
frustates rotational support. (This is, after all, the physical
mechanism to turbulent mass accretion.) 
For example, \cite{fromang05} found numerical evidence that MHD
turbulence can enhance GI and lead to disk fragmentation (see also
\S\ref{MHD}).

In this work, we use viscosity to model such a 
physical effect due to turbulence, but we emphasize that viscous 
GI does not assume a particular origin for the viscosity. We will
generalize the theory of viscous GI to include an energy equation  with viscous
heating, irradiation, explicit cooling, as well as
three-dimensionality (3D). This allows us to consider viscous GI in
more realistic PPD models.

@@

\subsection{Relevance to gravito-turbulent disk fragmentation}
We develop a general framework for viscous disks without  
assuming a specific origin of the turbulence that this viscosity 
represents. %without assuming the nature of the turbulence represented
            %by this viscosity  
We will, however, apply the theory to `gravito-turbulent'
disks in which the viscosity associated with some underlying (classic)
GI, as described below.

Consider a disk initially laminar without a heating mechanism and
cools. Then the disk temperature will be lowered until $Q=O(1)$, whence
non-axisymmetric modes grow and heats the disk through the dissipation
of spiral shocks. %new rafikov 16 
This permits a quasi-steady, turbulent state with
$Q =O(1)$ in which cooling is balanced by shock heating to maintain
thermal equilibrium \citep{gammie01}. 
Represent the dynamic and thermodynamic effect of this turbulence by a
viscosity. Then the $\alpha$ and $\beta$ parameters defined above will
be inversely related.

Experimentally, however, it is found that if the cooling rate or
viscosity is too large, say,   
\begin{align}\label{frag_cond}
  \beta < \beta_\mathrm{c} \quad
  \left(\alpha>\alpha_\mathrm{c}\right), 
\end{align}
then the disk fragments \citep{gammie01,rice05,rice11}. In fact, in
the absence of global effects, numerical simulations show there are
two --- and only two --- possible outcomes for self-gravitating,
cooling disks: gravito-turbulence or fragmentation.  
%more accurately: given a total simulation time, there are two
%outcomes 
%sufficiently long sims -> always frag?

There is considerable debate on the exact value of $\beta_\mathrm{c}$ 
and whether or not a critical cooling time can be defined at all 
\citep{meru11,lodato11,meru12,paardekooper12,hopkins13}. 
%mention numerical effects? 
However, it is generally accepted that \emph{steady, gravito-turbulent
  disks do not  
  exist for sufficiently rapid cooling or large viscosity}.  
This intriguing because, as highlighted in
\S\ref{cool_gi}---\ref{visc_gi}, these effects can reduce
gravitational stability. This motivates a physical    
interpretation of disk fragmentation as the inability to maintain a
gravito-turbulent state due to the action of cooling and/or viscosity.

>>>>

%\documentclass[12pt]{aastex}
\documentclass[iop, numberedappendix]{emulateapj}
%\documentclass{article}
%\usepackage{draftwatermark}
%\usepackage{paralist}
\shorttitle{Non-ideal Gravitational instability}
\shortauthors{M.-K.\ Lin \& K. M. Kratter}

@@

\begin{abstract}
Low mass, self-gravitating accretion disks admit quasi-steady,
`gravito-turbulent' states in which cooling balances turbulent {\bf viscous?}
heating. However, numerical simulations show that gravito-turbulence
cannot be sustained beyond dynamical timescales when the cooling rate is too fast,
or corresponding turbulent viscosity is too large. The result is is disk fragmentation.        
We motivate and quantify an interpretation of disk % a physical
                                % interpretation 
fragmentation as the inability to maintain gravito-turbulence due to  
{\bf proper / formal / etc?} secondary instabilities driven by:  
\begin{inparaenum}[1)] 
\item 
  cooling, which reduces pressure support; and/or
\item 
  viscosity, which reduces rotational support. 
\end{inparaenum}
We analyze the gravitational stability of accretion disks 
including these non-ideal effects explicitly.    
We derive a general dispersion relation for linear waves in a 
self-gravitating, viscous, non-adiabatic disk with internal heating,
external irradiation, and cooling. We  
consider parameterized cooling functions in two and three dimensions,
as well as %full 
radiative diffusion in 3D. We show that 
%cooling/viscosity can destabilize the disk even when the classic
%Toomre stability condition is satisfied; 
generally there is no critical cooling rate/viscosity below which 
the disk is formally stable, although interesting limits appear when we examine the growth
of modes of order the disk scale height. 
We apply this new linear theory to 
global models of protoplanetary disks (PPDs) subject to
gravito-turbulence modeled as an effective 
viscosity, and cooling via dust opacity. 
%We find the sole effect of
%cooling is insufficient to 
%destabilize a realistic PPD. 
We find that viscosity renders the PPD beyond $\sim 60$AU dynamically
unstable on radial %this is for Mdot = 1d-6  
lengthscales a few times the local disk thickness. This is coincident 
with the empirical condition for disk fragmentation based on a
maximum sustainable stress. 
We suggest turbulent stresses can play an active role in realistic disk 
fragmentation by removing rotational stabilization against 
self-gravity, and that the observed transition in behavior from gravito-turbulent to fragmenting may derive from
instability of the gravito-turbulent state itself.
\end{abstract}

\section{Introduction}\label{intro} 
Understanding the gravitational stability of rotating disks    
is central to many astrophysical problems. In the context of 
gaseous protoplanetary disks (PPDs), gravitational instability (GI)
has two applications. It can provide gravitational torques to
transport angular momentum outwards and thus enable mass accretion
\citep{armitage10,turner14}. GI may also lead to disk fragmentation,
which has been invoked to explain giant planet formation at large
radii \citep{boss97,helled14} %% KMK need an original citation, not just review.
Studying these non-linear phenomena requires
direct numerical simulations. Nevertheless, physical insight can be    
obtained through analytical modeling.

The standard metric for the (inverse) strength of disk self-gravity is
the \emph{Toomre parameter}, 
\begin{align}\label{toomreQ_criterion}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}  
\end{align}
\citep{toomre64}. Here, $c_s$ is the sound-speed,
$\kappa$ is the epicyclic frequency (which equals the rotation
frequency $\Omega$ in a Keplerian disk), $\Sigma$ is the surface
density and $G$ is the gravitational constant. The Toomre parameter is a
measure of the destabilizing effect of self-gravity ($G\Sigma$) against 
the stabilzing effect of rotation ($\kappa$) and pressure 
($c_s$). This is evident from the dispersion relation, 
\begin{align}\label{classic_dr}
  s^2 =  2\pi G\Sigma|k| - \kappa^2 - c_s^2k^2,   
\end{align}
which relates the growth rate $s$ and radial wavenumber $k$ for local,
axisymmetric waves in a two-dimensional (2D, razor-thin), inviscid and
adiabatic (including isothermal) disk. Such disturbances are formally
unstable when $Q<1$. Non-axisymmetric modes can develop for larger,
but still order-unity values of $Q$
\citep{lau_bertin78,papaloizou89,papaloizou91}. 
%strictly speaking non axi *local* modes are stable -> only transient
%growth
%true non-axi instability needs global disk  model 
%non-axi GI at slightly larger Q, but still characterized by Q

We emphasize that the oft-used dispersion relation and corresponding Toomre parameter, %It is important to remember that the popular dispersion relation, 
(Eqs. \ref{toomreQ_criterion} and \ref{classic_dr}) are derived from \emph{idealized} conditions: 
the base disk is laminar and does not experience any net thermal
losses\footnote{Isothermal disks implicitly assume  
  heating and thermal losses are exactly balanced at all 
  times, so these do not cool in the current context.} 
(hereafter `cooling').  
However, real disks can cool, for example, %radiative, convective cooling
due to radiative losses. Accretion disks may also be turbulent, the dynamic and thermodynamic effect of which is often modeled through an
effective viscosity \citep{shakura73,lin87,armitage01,rafikov15}. 
%{\bf  more refs here?} 
%in terms of AM transport and heating 
%assess GI in general

Despite this, the Toomre parameter %and its associated
%dispersion relation are
 is still widely applied to viscous (turbulent), cooling
accretion disk models \citep[e.g.][]{gammie01,cossins09, kimura12}, {\bf more refs?}
%%KMK
despite the mismatch in the included underlying physics.
%even though Eq. \ref{toomreQ_criterion}---\ref{classic_dr} were 
%derived from inviscid disks without cooling. 
%e.g. using classic DR to determine what scales need to be resolved

We show in this work that including non-ideal
physics, such as cooling and viscosity, modifies the classic dispersion
relation, and hence the condition for GI. Thus the Toomre condition
alone is insufficient to assess GI in realistic disks.  The goal of
this work is to generalize the analytic treatment of disk GI to allow
a self-consistent discussion of GI in realistic accretion disk
models. Specifically, we will examine how  cooling and viscosity can
lead to GI even when $Q>1$, by removing pressure and rotational
support, respectively.  We also review the origins for the transition from self-regulated GI 
to fragmentation seen in numerical simulations. {\bf add sentence here more like s 1.4 laying
out what you are about to talk about. at the very least a transition sentence or two?}

\subsection{Cooling-driven gravitational instability}\label{cool_gi}
Cooling reduces pressure support against self-gravity. The
cooling time  
$\tcool$ is the timescale over which the disk temperature $T$ is relaxed
to some floor value $T_0$, which may be zero. 
%That is, $\p_t T =
%-(T-T_0)/\tcool$. 
 It is often written as 
\begin{align} \label{beta_def}
  \tcool = \beta \Omega^{-1},
\end{align}
where $\beta$ is the corresponding dimensionless cooling time. This
type of parameterized cooling, first applied by \cite{gammie01},
allows a range of thermodynamic responses to be explored. In reality, PPD cooling is controlled by radiation 
from dust grains \citep{bell94,dalessio97,chiang97}. Although most PPDs subject to GI are optically thick, 
cooling parameterized by the $\beta$ model above only captures optically thin cooling when used in numerical
simulations.
%{\bf physical origin of cooling in PPDs}

%%KMK 
Previous work has quantified the role of cooling primarily as a means to reduce the sound speed term
in $Q$ (though see e.g. \citealt{clarke07}). Here we will quantify how cooling enables GI
even when $Q>1$. For example, if perturbations can
cool towards zero temperature, we find the above dispersion relation is
modified to read 
\begin{align}\label{cool_dr}
s^2 = 2\pi G\Sigma |k| - \kappa^2 - \gamma \left(\frac{\tcool s}{1 +
  \tcool s}\right)c_s^2k^2,
\end{align}
where $\gamma$ is the adiabatic index. Cooling increases the growth
rate by reducing the magnitude of the 
pressure term. In fact, for $t_c\in[0,\infty)$ a formal condition
for instability is  
\begin{align*}
  2\pi G \Sigma |k| > \kappa^2,
\end{align*}
which is what would be obtained from Eq. \ref{classic_dr} with 
pressure neglected ($c_s\to 0$). This condition does not actually
depend on the cooling rate, and instability is
possible for any finite $Q$. %at sufficiently small scales 
This shows that the mere presence of
cooling changes the qualitative nature of GI compared to the simple
Toomre condition.   
%{\bf numerical implications? if using standard beta cooling, then as
%  long as it's there then sufficiently small scales will be
 % unstbale. can access those scales with high resolution. defer to
 % discussion section}

@@

\subsection{Viscosity-driven gravitational instability}\label{visc_gi}
%there is also viscous damping which stabilizes
Viscous disks can also develop GI even when $Q>1$. This is because, as 
demonstrated below, viscosity removes rotational support against  
self-gravity for long-wavelength disturbances 
\citep{lynden-bell74,willerding92,gammie96}. A similar effect   
occurs in dusty fluids where the required frictional forces are 
provided by dust-gas drag \citep{ward00, takahashi14}. 
In fact, this
is a mechanism to enhance particle clumping for planetesimational
formation %frag of dust 
\citep{youdin11}. Therefore it is not unreasonable to expect analogous
fragmentation in gaseous disks due to viscosity.

It is conventional to write the kinematic viscosity $\nu$ as   
\begin{align}\label{alpha_param}
  \nu  = \alpha \frac{c_s^2}{\Omega}, %dimensional argument 
\end{align}
where $\alpha$ is the dimensionless viscosity
coefficient \cite{shakura73}. This
parameterization can be modified to include more complex dependencies on the
fluid variables, but Eq. \ref{alpha_param} is the general form. 
%KMK I think this is sufficient here {\bf maybe no need to intro alpha param here?}

For an isothermal, viscous, self-gravitating disk in 2D,
\cite{gammie96} finds the approximate dispersion relation   
\begin{align}
s \simeq \frac{\nu k^2\left(2\pi G\Sigma |k| -
  c_s^2k^2\right)}{\kappa^2 + c_s^2k^2 - 2\pi G\Sigma|k|}. 
\end{align}
Assuming $Q>1$, instability occurs if 
\begin{align*}
  2\pi G \Sigma > c_s^2|k|,
\end{align*}
provided that $\nu k^2\neq0$. This condition for viscous GI is 
identical to what would be obtained from Eq. \ref{classic_dr} with
rotation neglected ($\Omega\to 0$). That is, a classically stable
disk can be destabilized by viscosity as it effectively eliminates
rotational stabilization \citep{lynden-bell74}.

%turbulence <-> viscosity assumption 
For gaseous accretion disks, a Navier-Stokes viscosity is often
implemented (as above) to mimic hydrodynamic or magneto-hydrodynamic
(MHD) turbulence \citep{shakura73}. How well turbulence can be modeled
as an effective viscosity is a separate issue 
\citep{balbus99}. However, it is reasonable to assume that these mechanisms, which are observed in simulations 
to transport angular momentum outwards (and mass inwards),
frustate rotational support. %KMK (This is, after all, the physical mechanism to turbulent mass accretion.) 
%%KMK this comment seems out of place here 
%For example, \cite{fromang05} found numerical evidence that MHD
%turbulence can enhance GI and lead to disk fragmentation (see also
%\S\ref{MHD}).

In this work, we use viscosity to model two possible
physical effects of turbulence: heating via dissipation and angular momentum transport. 
%KMK
We emphasize that our model for GI in the presence of viscosity (hearafter Viscous GI)
does not assume a particular origin for the viscosity. GI itself or other forms of magnetic or hydrodynamic turbulence are allowed
in this framework. We will
generalize the theory of viscous GI to include an energy equation  with viscous
heating, irradiation, explicit cooling, as well as
three-dimensionality (3D) in order to consider viscous GI in
more realistic PPD models.

@@

\subsection{Relevance to gravito-turbulent disk fragmentation}
We develop a general framework for viscous disks without  
assuming a specific origin of the turbulence that this viscosity 
represents. %without assuming the nature of the turbulence represented
            %by this viscosity  
We will, however, apply the theory to `gravito-turbulent'
disks in which the viscosity is associated with some underlying (classic)
GI, as described below.

Consider an initially laminar, $Q>>1$ disk, without an external heating mechanism, as it
cools. The disk temperature will decline until $Q=O(1)$, whence
non-axisymmetric modes grow and heat the disk through the dissipation
of spiral shocks \citep{cossins09}. %new rafikov 16 
This setup permits a quasi-steady, turbulent state with
$Q =O(1)$ in which cooling is balanced by shock heating to maintain
thermal equilibrium \citep{gammie01}. 
One may represent the dynamic and thermodynamic effect of this turbulence by a
viscosity. Then the $\alpha$ and $\beta$ parameters defined above will
be inversely related ({\bf KMK see foreward ref eq...?}).

Numerical experiments, however, show that if the cooling rate or
viscosity is too large, say,   
\begin{align}\label{frag_cond}
  \beta < \beta_\mathrm{c} \quad
  \left(\alpha>\alpha_\mathrm{c}\right), 
\end{align}
then the disk fragments \citep{gammie01,rice05,rice11}. In fact, in
the absence of global effects, numerical simulations show that there are
two --- and only two --- possible outcomes for self-gravitating,
cooling disks: gravito-turbulence or fragmentation. %more accurately: given a total simulation time, there are two
%outcomes 
%sufficiently long sims -> always frag?

There is considerable debate on the exact value of $\beta_\mathrm{c}$ 
and whether or not a critical cooling time can be defined at all 
\citep{meru11,lodato11,meru12,paardekooper12,hopkins13}. {\bf KMK We defer a discussion of
numerical convergence issues to Section \ref{sec-disc}.?}
%mention numerical effects? 
However, it is generally accepted that \emph{steady, gravito-turbulent
  disks do not  
  exist for sufficiently rapid cooling or large viscosity}.  
This  is intriguing because, as highlighted in
\S\ref{cool_gi}---\ref{visc_gi}, these effects can reduce
gravitational stability independently of their influence on the exact value of the classic $Q$. This motivates a physical    
interpretation of disk fragmentation as the inability to maintain a
gravito-turbulent state due to %the action of 
%KMK
secondary instabilities driven by cooling and/or viscosity.
